---
title: "Lab 4: Factors"
author: "Environmental Data Analytics | John Fay and Luana Lima "
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---


```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
#Read packages
library(tidyverse);library(lubridate); library(here)

#Read in data, don't convert strings to factors
sites <- read.csv(here('Data/Raw/NWIS_SiteFlowData_NE_Raw.csv'))
dim(sites)

#Convert date column to date object
sites$date <- ymd_hms(sites$date)

```


## Examine the lakename column as a character column
* Tip - Lubridate cheatsheet: <https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf>
```{r create and explore days of week column}

#Compute the day of the week each sample was collected
sites$dow <- wday(sites$date)
#this is new colum 
#wday is a lubridate function that looks at day of year and returns day of the week


## What class is the dow column?
class(sites$dow)


#List the unique values in the column
unique(sites$dow)

#Summary - is it meaningful?
summary(sites$dow)
#not all numbers have numeric meaning (are ordinal, though)
#cannot take mean of this
```

## Convert the days of the week to a factor
```{r convert dow column to factor and explore}

#Convert to a factor
sites$dow <- factor(sites$dow)
#convert column from numeric factor to column

## What class is the dow column now?
class(sites$dow)

#List the unique values in the column
unique(sites$dow)
#shows unique values

#Summary - is it meaningful?
summary(sites$dow)

#Show the levels associated with our factor column
levels(sites$dow)
#set of unique values in dataset
#every row in dataset has a relationship - all of these collected on same day, for example (factor is a grouping variable)
#when would a string not be grouping variable - a dna sequence where letters mean something, would be an example
```

## Factors & Levels
A **factor** is a grouping variable: rows with the same value in the factor column are members of the same group. For example, all rows with a `dow` of 3 were collected on a Tuesday.

**Levels** are the unique values that the factors can take. The `unique()` function reveals the levels as does the `levels()` function. By default, the levels includes only the values in the factor column. However, we can add additional levels and we can also re-order levels. Why would we want to do this? Well, one example would be if we wanted to plot the number of samples recorded each day of the week, and in that plot we wanted to include all days of the week, not just the one's found in our data...

```{r plot samples recorded each day of the week}
#Create a bar plot showing the count of samples by day of week
ggplot(sites,aes(x=dow)) +
  geom_bar()
```



```{r add levels to our data}
#Update the levels to include all seven days of the week (note we have to start with existing levels)
levels(sites$dow) <- c(levels(sites$dow),"1","2","6","7")
#taking existing levels, and adding additional ones and completing the set

#Confirm the change
levels(sites$dow)

#Plot again
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)
#count of records in each of these levels
#plots it again using just standard bar plot statement. But next is visualization. - use scales, which override defaults. If you have levels that have no members, dont drop them
```

We now have all days of the week, but the order is incorrect. Also, we have day numbers, not day names. How to fix?
```{r reorder and rename levels}

#Reorder by overwriting the column with the existing values assigned with a new order of levels
sites$dow <- factor(sites$dow, levels=c("1","2","3","4","5","6","7")) #Or... as.character(seq(1,7))
#as.character(seq(1,7))
#updates column,
#control categories, not just the ones that have members
#using numbers currenttly as days of the week
#factors also have labels

#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)

#Add **labels** to our factors
sites$dow <- factor(
  sites$dow, 
  levels=c("1","2","3","4","5","6","7"),
  labels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat"))
#here are the levels, and here are the labels
#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)
```

## An exercise for you: see if you can plot the number of samples recorded each month