---
title: "Assignment 2: Coding Basics"
author: "Ellen Nirenblatt"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
thirty_seq <- seq(1, 30, 3) #from 1 to 30, by 3
thirty_seq # runs the function

#2. 

mean(thirty_seq)
median(thirty_seq)



#3. 

#make mean and median into objects
x <- mean(thirty_seq)
y <- median(thirty_seq)

#name function and compare values of mean(x) and median(y)
sequenceFunction <- function() {
  if(x > y) {
    print(FALSE)
  } 
  else if (x < y) {
    print(FALSE)
  }
  else {
    print(TRUE)
  }
}

sequenceFunction()
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5. 

students <- c("caroline","charles","ellen","bella") #vector type is a string of characters
scores <- c(82, 84, 93, 97) #vector type is an integer
passingscore <- (scores > 50) #vector type is Boolean

is.vector(students)
is.vector(scores)

testresults <- data.frame(
  Names <-students,
  testscores <-scores,
  Result <-passingscore
  
)

print(testresults)



```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:

#This data frame is different from a matrix because a matrix consists of the same data type. For example, a matrix would consist only of integers or characters. This data frame has three types of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#passinggrade <- function(x){
 # if (scores > 50) {
#    print("True")
    
#  }
#  else if (scores < 50) {
#    print ("false")
#  }
#  else {
#    print ("50")
 # }
#}
#passinggrade()

passedexam <- function() {
  ifelse(scores>50,TRUE,FALSE)
}
passedexam()
  



```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked and not the "if and else". I found that the reason was because the first "if, else" function only works when there is one variable in a list/it will only work on the first variable in a list. By using the ifelse statement, it iterates through each variable in the list, which is why it returned multiple values. 
