---
title: "Assignment 5: Data Visualization"
author: "Ellen Nirenblatt"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)
library(ggridges)
here()
getwd()
nutrients <- 
  read.csv(here("Assignments/Processed_KEY/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)
#nutrients
litter <- 
  read.csv(here("Assignments/Processed_KEY/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)
#litter




#2 

nutrients$sampledate <- ymd(nutrients$sampledate)
litter$collectDate <- ymd(litter$collectDate)

#nutrients
#litter

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme, fig.height = 3, fig.width = 4}


customtheme <- theme_light(base_size = 12) +
  theme(axis.text = element_text(color = "purple"), 
        legend.position = "left")
theme_set(customtheme)

graphnutrients <- 
  ggplot(nutrients) +
  geom_point(aes(x = depth, y = irradianceWater)) +
  customtheme
#print(nutrients)
  
#3


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

PvPo4 <- 
  ggplot(nutrients, aes(x= po4, y=tp_ug, color = lakename))+
  geom_point(aes(x = po4, y=tp_ug))+
  xlim(0,50)+
  ylim(0,100)+
  geom_smooth(method=lm, color= "black")
  
  
print(PvPo4)

  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
 #geom_point() +

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

boxtemp <- 
  ggplot(nutrients, aes(x= month, y=temperature_C, color = lakename))+
  geom_boxplot(aes(x = factor(month, levels= 1:12, labels = month.abb), y=temperature_C))
print(boxtemp)

boxTP <-
  ggplot(nutrients, aes(x= month, y= tp_ug, color = lakename))+
  geom_boxplot(aes(x = factor(month, levels= 1:12, labels = month.abb), y=tp_ug))
print(boxTP)

boxTN <- 
  ggplot(nutrients, aes(x= month, y= tn_ug, color = lakename))+
  geom_boxplot(aes(x = factor(month, levels= 1:12, labels = month.abb),y=tn_ug))
print(boxTN)

combinedplot <- plot_grid(boxtemp, boxTP, boxTN, x= month)
print(combinedplot, ncol = 2, align = "v")







  
                                                                                  

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
I observed that the temperatures decreased in later months (winter) and the TP and TN also followed that pattern.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

littersubset <- subset(litter, functionalGroup == "Needles")
#print(littersubset)

needleplot <- ggplot(littersubset, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  customtheme

print(needleplot)


#7


littersubset2 <- subset(litter, functionalGroup == "Needles")
#print(littersubset)

needleplot2 <- ggplot(littersubset, aes(x = collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(littersubset2$nlcdClass)
  customtheme

print(needleplot2)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
I think that the plot for 7 is more effective because it shows clearer patterns and separates out the classes more efficiently, and makes it easier to read. 